Verification of concurrent design patterns with data

### 2019-02-19 2018-Arch-pNets/arch-pNets.tex
arch-pNets paper: Fixed a bug in Fig 1 (fail |--> {zone}, instead of reset |--> {zone})
eric comments

\begin{abstract}
We provide a solution for the design of safe concurrent systems by
compositional application of verified design patterns\mdash which we
call \emph{architectures}\mdash to a \noteEM{ca veut dire quoi, minimal? On enleve ?}minimal set of functional
components.  To this end, we extend the theory of architectures
developed previously for the BIP framework with the elements necessary
for handling data: definition and operations on data domains, syntax
and semantics of composition operators involving data transfer.  Under
monotonicity assumptions on guards and assignments, composition of
architectures preserves their characteristic safety properties.
To verify that individual architectures do enforce their associated
properties, we provide an encoding into open pNets, an intermediate
model  that supports SMT-based verification.
Additionally, the translation into  open pNet  gives
us a strong insight and a comparison of the coordination mechanisms
provided by the two frameworks.  The approach is illustrated by a case
study based on a previously developed BIP model of a nanosatellite
on-board software.

\keywords{Verification, \noteEM{Je mettrais bien symbolic quelque part...}symbolic automaton, composition, safety, interaction
  models.}
\end{abstract}

@@

We propose   to encode BIP architectures
into open pNets and to verify, using SMT techniques,
the properties of the architectures. The encoding also allows us to
compare BIP architectures and  pNet formalism.  
Overall, the encoding allows us to precisely compare the coordination
mechanisms of the two models but also the way component composition is
expressed in the two formalisms. \noteEM{J'ai vire une phrase un peu
  redondante avec la phrase precedente, et deja une autre 20 lignes
  plus haut... }
%This also gives the opportunity
%to relate two  different visions on compositional verification.

@@

%\paragraph{Operational Semantics.}
%\label{section:op-semantics}
\renewcommand{\Pred}{g}
\renewcommand{\Post}{u}
The semantics of an open pNet is expressed as  an automaton where each transition \noteEM{remplacer composes par ``coordinates'' ?}composes the actions of several holes, 
the transition occurs if some predicates hold, and can involve  
 state modifications.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transition]
	\label{defn:OpenTransitions}
	An \emph{open transition} over a
	set $J$ of holes  and a set of states $\mathcal{S}$ is 
	a structure of the form:
\\[-2ex]	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{	\beta_j^{j\in J}, \Pred, \Post}
	{s \OTarrow {{\alpha}}s'}
	\end{mathpar}
	Where $s, s'\in\mathcal{S}$ and $\beta_j\in\actions{\variables}$
        is an action of the hole $j$. $\alpha$ is the resulting global action. \Pred\ is a predicate 
	over the different variables of the
	terms, labels, and states  $\beta_j$, $s$, $\alpha$. $\Post\in 
	\assigns{\variables}$\ is 
	a set of assignments that are the effects of the transition.
Open transitions are identified modulo logical equivalence on their predicate.
\end{definition}
%\begin{definition}[Open transitions]
%	\label{def:OpenTransitions}
%	An \emph{open transition} over a set \noteSB{In the architecture section, I say that 
%we omit indices on $\goesto{}$ and that they are always clear from the 
%context.}\hl{$(S_i,s_{0 i}, \rightarrow_i)^{i\in
%	I}$} of LTSs, a
%	set $J$ of holes with sorts $\Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is 
%	a structure of the form:	
%	\begin{mathpar}
%	\inferrule*[myfraction=\reddottedrule]
%	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
%		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred, \Post}
%	{s \OTarrow {v}s'}
%	\end{mathpar}
%	Where $s, s'\in\mathcal{S}$ and for all
%        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
%	LTS $(S_i,s_{0 i}, \rightarrow_i)$, and \noteSB{Why insist on ``transition'' and 
%$\goesto{b_j}_j$? Why not simply ``$b_j$ is an action in the sort 
%$\Sort_j$?}\hl{$\xrightarrow{b_j}_j$
%        is a transition of the hole $j$}, for any action $b_j$ in the
%        sort $\Sort_j$.  
%        of this open transition. \Pred\ is a predicate 
%	over the different variables of the
%	terms, labels, and states $s_i$, $b_j$, $s$, $v$. \Post\ is a set of equations that 
%	hold \emph{after the open transition}, they are represented as a substitution of the 
%	form $\{x_k\gets \Expr_k\}^{k\in K}$ \noteLH{considering the level of abstraction I 
%	would like to just put $\Post_=e$}
%	where $x_k$ are variables of $s'$, $s'_i$, and $e_k$ are expressions over the other 
%	variables of the open transition.
%\end{definition}

@@

\section{Related work}
\label{secn:related}
Previous works on pNets and BIP architectures have been mentioned at the beginning of the paper. Consequently, we  review here the closest related works outside the scope of pNets and BIP architectures. Among those, we focus on existing approaches for the compositional and parameterized verification of systems, and on alternative verification engines???
\noteLH{not sure about this classification ; to be refined}

@@

%%% compositional verification
approaches by typing interfaces/protocols == Simon

@@

Basic research on behaviour models and verification algorithms for
data-sensitive systems started in the nineties, with the seminal work
of Hennesy, Lin, and their colleagues on value-passing systems with
assignments
\cite{CONCUR::Lin1996,HennessyL1995:353,APSEC::Lin2001}. Later, many
different works addressing various classes of infinite-state systems
and/or paramaterized topologies have been published, using
combinations of approaches, often including predicate abstraction and
SMT satisfiability (see
e.g. \cite{BruniEtAl-Tiles-Concur2000,DBLP:conf/cade/GhilardiNRZ08,DBLP:journals/jsat/AlbertiGPRR12,CimattiEtAl-NUXMV-CAV2014,ChampionEtAl-Kind2-CAV2016}). With
respect to these, we use symbolic representations not only to get a
finite representation of infinite spaces, but also to express the
(data-sensitive) synchronisations with the environment, making our
models suitable for compositional verification.

Among these works, several have shown the capacity of the SMT engines
(either Z3 or Yikes) as servers for solving verification conditions of
the algorithms, for large case-studies
(e.g. \cite{DBLP:journals/corr/CimattiGMT13,DBLP:journals/corr/abs-1806-11459}).
%% SMT on an application not using component structure. is there a strong Z3 example == Eric

As compositional proof of safety is difficult, some approaches rely on theorem proving to ensure the safety of component operations. Coqots and Pycots\cite{BCDLM:CBSE2014} even allow to prove the safety of reconfiguration procedures which are known to be highly difficult to verify, and massively parametrised. unfortunately the approach relies on a high expertise and significant efforts from the component user as (s)he must prove manually the safety of the reconfiguration process. In this paper, we rely on automatic encoding into a SMT solver making the proof of properties much more automatic, but we cannot prove the safety of the reconfiguration procedure.

In~\cite{DDJO:JLAMP2012}, the authors propose a compositional proof system for distributed objects that is suitable for implementation within the KeY framework~\cite{BHS:Key2007} and uses a Hoare logic approach.
The approach relies on history invariant for each object, separating local reasoning for proving object invariant from global reasoning that relies on well-formedness of traces. Compared to this  approach, we do not  deal with complex history-based specifications because we use interaction specification and SMT-reasoning instead of Hoare logic. Also the correctness of the composition is given by the properties of architectures which allows for more flexibility than well-formedness of traces.

>>>>

\begin{abstract}
We provide a solution for the design of safe concurrent systems by
compositional application of verified design patterns\mdash 
called \emph{architectures}\mdash to a small set of functional
components.  To this end, we extend the theory of architectures
developed previously for the BIP framework with the elements necessary
for handling data: definition and operations on data domains, syntax
and semantics of composition operators involving data transfer.  Under
monotonicity assumptions on guards and assignments, composition of
architectures preserves their characteristic safety properties.
To verify that individual architectures do enforce their associated
properties, we provide an encoding into open pNets, an intermediate
model  that supports SMT-based verification.
Additionally, the translation into  open pNet  gives
us a strong insight and a comparison of the coordination mechanisms
provided by the two frameworks.  The approach is illustrated by a case
study based on a previously developed BIP model of a nanosatellite
on-board software.

\keywords{Symbolic verification, composition, safety, interaction
  models.}
\end{abstract}

@@

We propose   to encode BIP architectures
into open pNets and to verify, using SMT techniques,
the properties of the architectures. The encoding also allows us to
compare BIP architectures and  pNet formalism.  
Overall, the encoding allows us to precisely compare the coordination
mechanisms of the two models but also the way component composition is
expressed in the two formalisms. 
%This also gives the opportunity
%to relate two  different visions on compositional verification.

@@

%\paragraph{Operational Semantics.}
%\label{section:op-semantics}
\renewcommand{\Pred}{g}
\renewcommand{\Post}{u}
The semantics of an open pNet is expressed as  an automaton where each transition coordinates the actions of several holes, 
the transition occurs if some predicates hold, and can involve  
 state modifications.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transition]
	\label{defn:OpenTransitions}
	An \emph{open transition} over a
	set $J$ of holes  and a set of states $\mathcal{S}$ is 
	a structure of the form:
\\[-2ex]	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{	\beta_j^{j\in J}, \Pred, \Post}
	{s \OTarrow {{\alpha}}s'}
	\end{mathpar}
	Where $s, s'\in\mathcal{S}$ and $\beta_j\in\actions{\variables}$
        is an action of the hole $j$. $\alpha$ is the resulting global action. \Pred\ is a predicate 
	over the different variables of the
	terms, labels, and states  $\beta_j$, $s$, $\alpha$. $\Post\in 
	\assigns{\variables}$\ is 
	a set of assignments that are the effects of the transition.
Open transitions are identified modulo logical equivalence on their predicate.
\end{definition}
%\begin{definition}[Open transitions]
%	\label{def:OpenTransitions}
%	An \emph{open transition} over a set \noteSB{In the architecture section, I say that 
%we omit indices on $\goesto{}$ and that they are always clear from the 
%context.}\hl{$(S_i,s_{0 i}, \rightarrow_i)^{i\in
%	I}$} of LTSs, a
%	set $J$ of holes with sorts $\Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is 
%	a structure of the form:	
%	\begin{mathpar}
%	\inferrule*[myfraction=\reddottedrule]
%	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
%		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred, \Post}
%	{s \OTarrow {v}s'}
%	\end{mathpar}
%	Where $s, s'\in\mathcal{S}$ and for all
%        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
%	LTS $(S_i,s_{0 i}, \rightarrow_i)$, and \noteSB{Why insist on ``transition'' and 
%$\goesto{b_j}_j$? Why not simply ``$b_j$ is an action in the sort 
%$\Sort_j$?}\hl{$\xrightarrow{b_j}_j$
%        is a transition of the hole $j$}, for any action $b_j$ in the
%        sort $\Sort_j$.  
%        of this open transition. \Pred\ is a predicate 
%	over the different variables of the
%	terms, labels, and states $s_i$, $b_j$, $s$, $v$. \Post\ is a set of equations that 
%	hold \emph{after the open transition}, they are represented as a substitution of the 
%	form $\{x_k\gets \Expr_k\}^{k\in K}$ \noteLH{considering the level of abstraction I 
%	would like to just put $\Post_=e$}
%	where $x_k$ are variables of $s'$, $s'_i$, and $e_k$ are expressions over the other 
%	variables of the open transition.
%\end{definition}

@@

\section{Related work}
\label{secn:related}
Previous works on pNets and BIP architectures have been mentioned at the beginning of the paper. We  review here the closest related works outside the scope of pNets and BIP architectures, with a particular focus on compositional and parameterized verification.

@@

%%% compositional verification
%approaches by typing interfaces/protocols == Simon

@@

Basic research on behaviour models and verification algorithms for
data-sensitive systems started in the nineties, with the seminal work
of Hennesy, Lin, and their colleagues on value-passing systems with
assignments
\cite{CONCUR::Lin1996,HennessyL1995:353,APSEC::Lin2001}. Later, many
different works addressing various classes of infinite-state systems
and/or paramaterized topologies have been published, using
combinations of approaches, often including predicate abstraction and
SMT satisfiability (see
e.g. \cite{BruniEtAl-Tiles-Concur2000,DBLP:conf/cade/GhilardiNRZ08,DBLP:journals/jsat/AlbertiGPRR12,CimattiEtAl-NUXMV-CAV2014,ChampionEtAl-Kind2-CAV2016}). With
respect to these, we use symbolic representations not only to get a
finite representation of infinite spaces, but also to express the
(data-sensitive) synchronisations with the environment, making our
models suitable for compositional verification.
Among these works, several have shown the capacity of the SMT engines
(either Z3 or Yikes) as servers for solving verification conditions of
the algorithms, for large case-studies
(e.g. \cite{DBLP:journals/corr/CimattiGMT13,DBLP:journals/corr/abs-1806-11459}). Also 
%% SMT on an application not using component structure. is there a strong Z3 example == Eric

As compositional proof of safety is difficult, some approaches rely on theorem proving to ensure the safety of component operations. Coqots and Pycots~\cite{BCDLM:CBSE2014} even manage to prove the safety of reconfiguration procedures which are known to be highly difficult to verify, and massively parametrised. The approach relies on a high expertise and significant efforts from the user. Here, we rely on automatic verification thanks to the SMT solver but we cannot prove the safety of the reconfiguration procedure.

In~\cite{DDJO:JLAMP2012}, the authors propose a compositional proof system for distributed objects that is suitable for implementation within the KeY framework~\cite{BHS:Key2007} and uses a Hoare logic approach.
%The approach relies on history invariant for each object, separating local reasoning for proving object invariant from global reasoning that relies on well-formedness of traces. 
Compared to this  approach, we do not  deal with complex history-based specifications and use interaction specification and SMT-reasoning instead of Hoare logic. 
%Also the correctness of the composition is given by the properties of architectures which allows for more flexibility than well-formedness of traces.
